---
title: "MALTEM Economics"
subtitle: "A brief overview"
author: "Laia Cirera, Joe Brew, Elisa Sicuri"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
fig_height: 5
fig_width: 7
output: ioslides_presentation

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, 
               echo = FALSE, 
               warning = FALSE, 
               message = FALSE, 
               error = TRUE, 
               cache = FALSE)
```

```{r, results = 'hide'}
# Packages
library(ggplot2)
library(cism)
library(rworldmap)
library(rgeos)
library(maptools)
library(rgdal)
library(tidyr)
library(RColorBrewer)
library(dplyr)
library(leaflet)
library(readr)
# Get world map
world <- readOGR('shp/world', 'ne_110m_admin_0_countries')



```

## Location

```{r}
par(mfrow = c(1,3))
cols <- ifelse(world$iso_a3 == 'MOZ', 'darkgreen', 'black')
plot(world,
     col = adjustcolor(cols, alpha.f = 0.5),
     lwd = 0.1,
     ylim = c(-5, 15),
     xlim = c(-15, 40))
title(main = 'Mozambique')

cols <- ifelse(moz2@data$NAME_1 == 'Maputo', 'darkgreen', 'black')
plot(moz2,
     col = adjustcolor(cols, alpha.f = 0.5),
     lwd = 0.1)
title(main = 'Maputo province')


map <- moz2[moz2@data$NAME_1 == 'Maputo',]
cols <- ifelse(map@data$NAME_2 == 'Magude', 'darkgreen', 'black')

plot(map,
     col = adjustcolor(cols, alpha.f = 0.5),
     lwd = 0.1)
title(main = 'Magude district')
par(mfrow = c(1,1))
```

## Questions

```{r, fig.height = 2}
dates <- seq(as.Date('2014-01-01'),
             as.Date('2016-12-31'),
             by = 1)
df <- data_frame(date = dates,
                 event = NA)

df$event[df$date == '2015-01-01'] <- 'Census'
df$event[df$date == '2015-08-01'] <- 'MDA'
df$event[df$date == '2016-01-01'] <- 'MDA'
df$event[df$date == '2016-12-15'] <- 'MDA'

ggplot(data = df,
       aes(x = date,
           y = 1,
           label = event)) +
  geom_label(nudge_x = 50) +
  theme_cism() +
  labs(x = 'Date',
       title = 'Intervention timeline') +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
      geom_vline(xintercept = as.numeric(df$date[!is.na(df$event)]),
               alpha = 0.6,
               lty = 3) 

```

1. How much did the intervention cost?

2. How much did the intervention save in medical costs?

3. How much did the intervention affect the "economic activity" (work and school productivity and performance) of the target population?

4. What will the affect of the intervention be across time (in the future) and space (in other places)?

## Xinavane

```{r}
map <- moz2[moz2@data$NAME_2 %in% c('Magude', 'Manhiça'),]
cism_map_interactive(lat = -25.053186, lng = 32.799852,
                     opacity = 0.8,
                     popup = 'Xinavane',
         # fspdf = moz2_fortified %>%
         #   filter(id %in% c('Magude', 'Manhiça')),
         point_size = 5) %>%
  addPolygons(data = map,
              color = c('red', 'blue'),
              popup = c('Magude', 'Manhiça'))
```

## Xinavane

```{r}
# Get info
load('../xinavane/data/read_in_finished.RData')
```


Absenteeism data spans from `r format(min(ab$date, na.rm = TRUE), '%B %d, %Y')` until `r format(max(ab$date, na.rm = TRUE), '%B %d, %Y')`. On average, there are `r ab %>% group_by(date) %>% tally %>% summarise(mean(n)) %>% as.numeric() %>% round` absences per day.


* Number of workers: `r length(unique(workers$number))` (ever)
  * 7,648 permanent
  * 29,872 temporary
* Average absenteeism rate: `r round(mean(df$absenteeism_rate), digits = 2)`% (data quality issues)


```{r}
temp <- ab %>%
  group_by(date) %>%
  tally
ggplot(data = temp, 
       aes(x = date, y = n)) +
  # geom_line(alpha = 0.6) +
  geom_point(alpha = 0.1) +
  geom_smooth() +
  xlab('Date') +
  ylab('Absences') +
  ggtitle('Crude worker absences') +
  theme_cism()
```


## Xinavane


```{r}

# Make "date" be the first day of the month
df$date <- df$month_start


temp <- df %>%
  group_by(date) %>%
  summarise(n = sum(eligibles),
            absences = sum(absences)) %>%
  mutate( p = absences / n * 100)
temp_g <- gather(temp, key, value, n:absences)
temp_g$key <-
  ifelse(temp_g$key == 'absences', 'Worker absence',
         ifelse(temp_g$key == 'n', 'Eligible worker-day',
                NA))
temp_g$key <-
  factor(temp_g$key, levels = rev(unique(temp_g$key)))

# Add a line for mda
temp_g$mda <- temp_g$date >= as.Date('2016-01-18')
temp$mda <- temp$date >= as.Date('2016-01-18')
# ggplot(data = temp_g,
#        aes(x = date, y = value, group = key, fill = key)) +
#   geom_area() +
#   xlab('Date') +
#   ylab('Number of eligible workers') +
#   ggtitle('Worker scheduling')

ggplot(data = temp) +
  geom_area(aes(x = date, y = n), fill = 'darkblue', alpha = 0.7) +
  geom_area(aes(x = date, y = absences), fill = 'darkorange', alpha = 0.6) +
  xlab('Date') +
  ylab('Worker-days') +
  labs(title = 'Absences (orange) and worker-days observed (blue)',
       subtitle = '(aggregated to monthly level)') +
  theme_cism()
```


## Xinavane


```{r}

# Make an mda dataframe for labeling
mda <- data.frame(date = as.Date('2015-10-01'),
                  mda = 'MDA',
                  y = 8)

ggplot(data = temp,
       aes(x = date, y = p)) +
  geom_area(alpha = 0.6, fill = 'darkorange') +
  xlab('Date') +
  ylab('Absenteeism rate (%)') +
  labs(title = 'Absenteeism rate during period of observation',
          subtitle = '(all causes)') +
  # geom_smooth() +
  geom_vline(data = mda, aes(xintercept = as.numeric(date)),
             alpha = 0.7,
             lty = 3) +
  geom_label(data = mda,
             aes(x = date,
                 y = y,
                 label = mda)) +
  theme_cism()

```


## Xinavane

```{r}

temp <- df %>%
  group_by(date) %>%
  summarise(n = sum(eligibles),
            absences = sum(sick_absences)) %>%
  mutate( p = absences / n * 100)
temp_g <- gather(temp, key, value, n:absences)
temp_g$key <-
  ifelse(temp_g$key == 'absences', 'Worker absence',
         ifelse(temp_g$key == 'n', 'Eligible worker-day',
                NA))
temp_g$key <-
  factor(temp_g$key, levels = rev(unique(temp_g$key)))

# Add a line for mda
temp_g$mda <- temp_g$date >= as.Date('2016-01-18')
temp$mda <- temp$date >= as.Date('2016-01-18')

ggplot(data = temp,
       aes(x = date, y = p)) +
  geom_area(alpha = 0.6, fill = 'darkorange') +
  xlab('Date') +
  ylab('Absenteeism rate (%)') +
  labs(title = 'Absenteeism rate during period of observation',
          subtitle = '(marked as sickness only)') +
  # geom_smooth() +
  geom_vline(data = mda, aes(xintercept = as.numeric(date)),
             alpha = 0.7,
             lty = 3) +
  geom_label(data = mda,
             aes(x = date,
                 y = y,
                 label = mda)) +
  theme_cism()
```


## Xinavane


```{r}
temp <- df %>%
  group_by(date, worker_type) %>%
  summarise(n = sum(eligibles),
            absences = sum(absences)) %>%
  mutate( p = absences / n * 100)

# Add a line for mda
temp$mda <- temp$date >= as.Date('2016-01-18')

cols <- brewer.pal(n = length(unique(temp$worker_type)), 'Spectral')

ggplot(data = temp,
       aes(x = date, y = p)) +
  geom_line(alpha = 0.3, aes(group = worker_type, color = worker_type)) +
  geom_point(alpha = 0.3, aes(group = worker_type, color = worker_type)) +
  geom_smooth(alpha = 0.1, aes(group = worker_type, color = worker_type)) +
  xlab('Date') +
  ylab('Absenteeism rate') +
  scale_color_manual(values = cols,
                     name = 'Worker type') +
  ggtitle('Worker absenteeism by type') +
    geom_vline(data = mda, aes(xintercept = as.numeric(date)),
             alpha = 0.7,
             lty = 3) +
  geom_label(data = mda,
             aes(x = date,
                 y = y,
                 label = mda)) +
  theme_cism()
```


## Xinavane


```{r}

did <- 
  df %>% 
  filter(month_start >= '2015-04-01',
         !is.na(location_laia)) %>%
  mutate(mda = ifelse(month_start >= '2015-12-01', 'Post', 'Pre')) %>%
  group_by(location_laia, mda) %>%
  summarise(absences = sum(absences, na.rm = TRUE),
            sick_absences = sum(sick_absences, na.rm = TRUE),
            eligibles = sum(eligibles, na.rm = TRUE)) %>%
  mutate(absenteeism_rate = absences / eligibles * 100,
         sickness_absenteeism_rate = sick_absences / eligibles * 100) %>%
  mutate(mda_num = ifelse(mda == 'Pre', 1, 2))

did$mda <- factor(did$mda, levels = c('Pre', 'Post'))

cols <- colorRampPalette(brewer.pal(n = 9, 'Set1'))(length(unique(did$location_laia)))

cols <- colorRampPalette(brewer.pal(n = 9, 'Set1'))(length(unique(did$location_laia)))
ggplot(data = did,
       aes(x = mda,
           y = absenteeism_rate,
           color = location_laia,
           group = location_laia)) +
  geom_line() +
  geom_point() +
  geom_label(aes(label = round(absenteeism_rate, digits = 3)),
             alpha = 0.5) +
  theme_cism() +
  xlab('Time') +
  ylab('Absenteeism rate') +
  ggtitle('All absenteeism rate: before and after',
          'Using company-based residential location data') +
  scale_color_manual(name = 'Location',
                     values = cols)
```


## Xinavane

```{r}


# Standardize scale to start
did <- did %>%
  arrange(mda) %>%
  group_by(location_laia) %>%
  mutate(standardized = absenteeism_rate / first(absenteeism_rate) * 100)

ggplot(data = did,
       aes(x = mda,
           y = standardized,
           color = location_laia,
           group = location_laia)) +
  geom_line() +
  geom_point() +
  geom_label(aes(label = round(standardized, digits = 3))) +
  theme_cism() +
  xlab('Time') +
  ylab('Absenteeism rate') +
  ggtitle('Relative scaled all absenteeism rate: before and after',
          'Using company-based residential location data') +
  scale_color_manual(name = 'Location',
                     values = cols)
```

## Schools

```{r}
source('../maltem_school_absenteeism/read_and_clean_all.R',chdir = TRUE)



locations <- schools %>%
  group_by(SCHOOL_NAME) %>%
  summarise(lng = mean(lng),
            lat = mean(lat))
locations <- locations %>%
  filter(SCHOOL_NAME %in% c('EPC 3 de Fevereiro',
                            'EPC Duco',
                            'EPC Graca Machel',
                            'EPC Ilha Josinha Machel',
                            'EPC Magude',
                            'EPC Maguiguana',
                            'EPC Maragra',
                            'EPC Moine',
                            'EPC Panjane',
                            'EPC Xinvane',
                            'EPC Simbe'))
locations$lng[locations$SCHOOL_NAME == 'EPC Magude'] <-
  32.799852
locations$lat[locations$SCHOOL_NAME == 'EPC Magude'] <-
  -25.05319

map <- moz2[moz2@data$NAME_2 %in% c('Magude', 'Manhiça'),]
cism_map_interactive(lat = locations$lat, 
                     lng = locations$lng,
                     opacity = 0.9,
                     popup = locations$SCHOOL_NAME,
         # fspdf = moz2_fortified %>%
         #   filter(id %in% c('Magude', 'Manhiça')),
         point_size = 5) %>%
  addPolygons(data = map,
              color = c('red', 'blue'),
              popup = c('Magude', 'Manhiça'))
```


## Schools

```{r}
source('../maltem_schools/monitor.R', chdir = TRUE)
df$id <- df$`Study Subject ID`
```

* Student-years: `r length(unique(df$id))`
* Student-days: `r nrow(df)` (only counting eligible school days)
* Number of students: `r length(unique(df$dob_E1_C1))` (and growing)
* 9 schools
* Average absenteeism rate of 6.2%



## Schools


```{r}

#####################################
# Visualize, etc.

# Students per school
x <- df %>%
  group_by(school,
           year) %>%
  summarise(n_students = length(unique(`Study Subject ID`))) %>%
  arrange(desc(n_students))

ggplot(data = x,
       aes(x = school,
           y = n_students)) +
  geom_bar(stat = 'identity',
           alpha = 0.6,
           fill = 'darkblue') +
  facet_wrap(~year) +
  theme_cism() +
  theme(axis.text.x = element_text(angle = 90)) +
  geom_label(aes(label = n_students)) +
  xlab('School') +
  ylab('Students') +
  ggtitle('Number of students observed by year/school')
```

## Schools

```{r}

# Absenteeism by date by intervention
x <- df %>%
  group_by(district, date) %>%
  summarise(absences = length(which(absent)),
            eligibles = n()) %>%
  ungroup %>%
  mutate(absenteeism_rate = absences / eligibles * 100)

library(RColorBrewer)
cols <- colorRampPalette(brewer.pal(n = 9, name = 'Set1'))(length(unique(x$district)))
ggplot(data = x,
       aes(x = date,
           y = absenteeism_rate,
           group = district,
           color = district)) +
  geom_point(alpha = 0.3,
             aes(size = eligibles)) +
  geom_smooth() +
  theme_cism() +
  scale_color_manual(name = 'District',
                     values = cols) +
  ylim(0, 25)
```

## Schools


```{r}
# Monthly absenteeism in both districts
x <- df %>%
  group_by(district, 
           date = month) %>%
  summarise(absences = length(which(absent)),
            eligibles = n()) %>%
  ungroup %>%
  mutate(absenteeism_rate = absences / eligibles * 100)

ggplot(data = x,
       aes(x = date,
           y = absenteeism_rate,
           group = district,
           fill = district)) +
  geom_bar(stat = 'identity',
           pos = 'dodge',
           alpha = 0.6) +
  theme_cism() +
  labs(title = 'Absenteeism rate') +
  scale_fill_manual(name = 'District',
                    values = cols) 
```

## Schools


```{r}
# Monthly absenteeism
x <- df %>%
  group_by(district, date = month) %>%
  summarise(absences = length(which(absent)),
            eligibles = n(),
            `Sample size` = length(unique(`Study Subject ID`))) %>%
  ungroup %>%
  mutate(absenteeism_rate = absences / eligibles * 100) 

ggplot(data = x,
       aes(x = date,
           y = absenteeism_rate)) +
  geom_bar(stat = 'identity',
           fill = 'darkorange',
           alpha = 0.6) +
  theme_cism() +
  labs(title = 'Absenteeism by month') +
  facet_wrap(~district, ncol = 2) +
  geom_hline(yintercept = mean(x$absenteeism_rate), lty = 2, alpha = 0.5)
```


## Schools



```{r}

# Lines instead of bars
g <- ggplot(data = x,
       aes(x = date,
           y = absenteeism_rate,
           color = district)) +
  geom_line(alpha = 0.5) +
  stat_smooth(aes(weight = `Sample size`), fill = NA) +
  # geom_point()
  geom_point(aes(size = `Sample size`),
             alpha = 0.3) +
  theme_cism() +
  labs(title = 'Absenteeism by month') +
  geom_hline(yintercept = mean(x$absenteeism_rate), lty = 2, alpha = 0.5) +
  scale_color_manual(name = 'District',
                     values = cols) +
  xlab('Date') +
  ylab('Monthly absenteeism rate') +
  geom_vline(xintercept = '2016-10-01')
g
```


## Schools



```{r}
# Examine by term
x <- df %>%
  group_by(district, year, term) %>%
  summarise(absences = length(which(absent)),
            eligibles = n(),
            `Sample size` = length(unique(`Study Subject ID`))) %>%
  ungroup %>%
  mutate(absenteeism_rate = absences / eligibles * 100)  %>%
  filter(!is.na(term)) %>%
  mutate(year_term = paste0(year, '-', term))

ggplot(data = x,
       aes(x = year_term,
           y = absenteeism_rate,
           color = district,
           group = district)) +
  geom_point() +
  geom_line()  +
  geom_vline(xintercept = 3.1, lty = 2, alpha = 0.6) +
  theme_cism() +
  xlab('Year-term') +
  ylab('Absenteeism rate') +
  scale_color_manual(name = 'District',
                     values = cols)
```


## Schools



```{r}
# Aggregate by year
x <- df %>%
  group_by(date = year, 
           district) %>%
  summarise(absences = length(which(absent)),
            eligibles = n()) %>%
  ungroup %>%
  mutate(absenteeism_rate = absences / eligibles * 100) 


ggplot(data = x,
       aes(x = factor(date),
           y = absenteeism_rate,
           color = district,
           group = district)) +
  geom_line() +
  geom_point() +
  theme_cism() +
  scale_color_manual(name = 'District',
                     values = cols) +
  ylim(0, max(x$absenteeism_rate)) +
  xlab('Date') +
  ylab('Absenteeism rate')
```

## Effectiveness

### Rainfall data

```{r}
source('../maltem_cost_effectiveness/master.R', chdir = TRUE)

ggplot(data = df,
       aes(x = date,
           y = precipitation)) +
  geom_line(alpha = 0.8,
            color = 'darkgreen') +
  facet_wrap(~district) +
  labs(x = 'Date',
       y = 'Milimeters',
       title = 'Rainfall',
       subtitle = 'Province of Maputo') +
  theme_cism()
```


## Effectiveness

### Rainfall data

```{r}

cols <- colorRampPalette(brewer.pal(n = 9, 'Spectral'))(length(unique(df$year)))
ggplot(data = df %>%
         mutate(day_of_year = as.numeric(format(date, '%j'))),
       aes(x = day_of_year,
           y = precipitation)) +
  geom_line(alpha = 0.8,
            aes(color = factor(year))) +
  facet_wrap(~district) +
  labs(x = 'Day of year',
       y = 'Milimeters',
       title = 'Rainfall',
       subtitle = 'Province of Maputo') +
  theme_bw() +
  scale_color_manual(name = 'Year',
                     values = cols)

```


## Effectiveness

### Rainfall data
```{r}

# Plot malaria risk
ggplot(data = df,
       aes(x = date,
           y = malaria_risk)) +
  geom_line(alpha = 0.3) +
  facet_wrap(~district) +
  xlab('Date') +
  ylab('Malaria risk') +
  ggtitle('Estimated climatological risk of malaria',
          'By district') +
  geom_line(data = df %>%
              group_by(district) %>%
              mutate(p = p / max(p)) %>%
              ungroup,
            aes(x = date,
                y = p),
            color = 'red',
            alpha = 0.5)


# Effectiveness: 
# BES evolution by district
# OPD evolution (Manhiça and Magude)

```

## Effectiveness

### IRS data

```{r}
# IRS coverage
x <- df %>%
  group_by(district, year) %>%
  summarise(irs_coverage = first(irs_coverage))

ggplot(data = x,
       aes(x = year,
           y = irs_coverage)) +
  geom_bar(stat = 'identity',
           pos = 'dodge',
           fill = 'darkorange', alpha = 0.6) +
  facet_wrap(~district) +
  theme_cism() +
  xlab('Year') +
  ylab('IRS coverage (%) of population') +
  geom_label(aes(label = paste0(round(irs_coverage, digits = 1), '%')),
             size = 3,
             alpha = 0.6) +
  labs(title = 'IRS coverage in Maputo province by district',
       subtitle = 'Number of people considered protected as % of total population')

```

## Effectiveness

### Outcomes

```{r}
ggplot(data = df,
       aes(x = date,
           y = pk)) +
  geom_line(aes(color = age)) +
  facet_wrap(~district) +
  labs(x = 'Date',
       y = 'Incidence (cases per 1,000)',
       title = 'Malaria incidence over time by district',
       subtitle = 'Cases per 1,000 inhabitants') +
  scale_color_manual(name = 'Age',
                     values = c('darkgreen', 'darkorange')) +
  theme_cism()
```

## Effectiveness

### Outcomes

```{r}
x <- df %>%
  mutate(magude = ifelse(district == 'MAGUDE', 'Magude', 'Not Magude')) %>%
    filter(district != 'NAMAACHA') %>%
  # filter(district != 'MANHICA') %>%
  group_by(date,
           # year, month,
           # age, 
           magude) %>%
  summarise(#pk = weighted.mean(pk, w = population),
            cases = sum(cases),
            pop = sum(population),
            precip = mean(precip)) %>%
  ungroup %>%
  mutate(p = cases / pop * 1000) #%>%
  # mutate(date = as.Date(paste0(year, '-',
  #                              '06',
  #                              # month, 
  #                              '-01')))# %>%
  # filter(date >= '2014-07-01')

library(ggthemes)
library(cism)
ggplot(data = x,
       aes(x = date,
           color = magude,
           y = p)) +
  geom_line(alpha = 0.6) +
  # geom_smooth() +
  # facet_wrap(~age, ncol = 1) +
  # theme_bw()  +
  scale_color_manual(name = '',
                     values = c('blue', 'red')) +
  # theme_fivethirtyeight() +
  geom_vline(xintercept = as.numeric(as.Date(paste0(2010:2017, '-12-31'))),
             lty = 2, alpha = 0.2) +
  geom_vline(xintercept = as.numeric(as.Date(paste0('2015-03-15'))),
             color = 'darkgreen') +
  labs(x = 'Week',
       y = 'Weekly incidence per 1,000',
       title = 'Weekly clinical malaria',
       subtitle = 'Province of Maputo, all ages, weekly') +
  geom_label(data = data_frame(x = as.Date(paste0('2015-03-15')),
                               y = 15,
                               label = 'MDA'),
             aes(x = x,
                 y = y,
                 label = label),
             color = 'black') +
  theme_cism()
```


## Effectiveness

### Outcomes


```{r}
ggplot(data = x %>%
         filter(date >= '2013-01-01'),
       aes(x = date,
           color = magude,
           y = p)) +
  geom_line(alpha = 0.6) +
  # geom_smooth() +
  # facet_wrap(~age, ncol = 1) +
  # theme_bw()  +
  scale_color_manual(name = '',
                     values = c('blue', 'red')) +
  # theme_fivethirtyeight() +
  geom_vline(xintercept = as.numeric(as.Date(paste0(2010:2017, '-12-31'))),
             lty = 2, alpha = 0.2) +
  geom_vline(xintercept = as.numeric(as.Date(paste0('2015-03-15'))),
             color = 'darkgreen') +
  labs(x = 'Week',
       y = 'Weekly incidence per 1,000',
       title = 'Zoom-in: Weekly clinical malaria',
       subtitle = 'Province of Maputo, all ages, weekly') +
  geom_label(data = data_frame(x = as.Date(paste0('2015-03-15')),
                               y = 15,
                               label = 'MDA'),
             aes(x = x,
                 y = y,
                 label = label),
             color = 'black') +
  theme_cism()
```

## Effectiveness

### Validation with OPD data

```{r}
source('../bes_vs_opd/compare.R', chdir = TRUE)

ggplot(data = master,
       aes(x = date,
           y = p,
           group = group,
           color = group)) +
  geom_point(alpha = 0.6) +
  geom_line(alpha = 0.3) +
  scale_color_manual(name = 'Source',
                     values = c('darkred',
                                'darkorange',
                                'darkgreen',
                                'green')) +
  facet_wrap(~district, nrow = 2) +
  theme_cism() +
  ylab('Percent of 2012 average') +
  xlab('Date') +
  ggtitle('BES and OPD data',
          'Fellow travelers')
```